---
title: "Factor Structure of the Hierarchial Personality Inventory for Children (HiPIC) as it applies to Autistic Chidlren"
author: "Straus, J., Burnheimer, E., Yo, J."
date: "June 8, 2021"
output: beamer_presentation
---

```{r echo=FALSE, results='hide', message=FALSE}

##############################
## libraries
##############################

library(readxl)
library(psych)
library(knitr)
library(GPArotation)
library(tidyverse)
library(qgraph)
library(polycor)
library(lavaan)
library(ltm)

##############################
## Read in Data
##############################

#Read in the data - this is for Jeffrey
hipic_data <- read_xlsx("C:/Users/yjeff/Box/255C Final Project/HiPIC raw data.xlsx")

#for all - run this to get rid of FamilyID
hipic_data <- subset(hipic_data, select=-FAMILYID)

##############################
## Recode Data
##############################

#Recoding Function

recode_d<-function(item) {
  item<-item
  original<-item
  item[item == 5] <- 99
  item[item == 4] <- 88
  item[item == 3] <- 3
  item[item == 2] <- 4
  item[item == 1] <- 5
  item[item == 99] <- 1
  item[item == 88] <- 2
  print(table(item))
  print(table(original))
  return(item)
}

# Recode Variables using Dr. Wood's SPSS syntax to identify the tables
# See "itemslist" rmarkdown file

# Create list of variables names that need to be recoded
var_names<-c( #Extraversion
            "PRE_PARENT_HIPIC_1", "PRE_PARENT_HIPIC_3","PRE_PARENT_HIPIC_4",
            "PRE_PARENT_HIPIC_11","PRE_PARENT_HIPIC_12", "PRE_PARENT_HIPIC_29",
             #Agreeableness
             "PRE_PARENT_HIPIC_34", "PRE_PARENT_HIPIC_39","PRE_PARENT_HIPIC_45",
            "PRE_PARENT_HIPIC_53","PRE_PARENT_HIPIC_55", "PRE_PARENT_HIPIC_56",
            "PRE_PARENT_HIPIC_57", "PRE_PARENT_HIPIC_63",
            #Conscientiousness
            "PRE_PARENT_HIPIC_80", "PRE_PARENT_HIPIC_81","PRE_PARENT_HIPIC_83",
            "PRE_PARENT_HIPIC_84","PRE_PARENT_HIPIC_85", "PRE_PARENT_HIPIC_91",
            "PRE_PARENT_HIPIC_94", "PRE_PARENT_HIPIC_95","PRE_PARENT_HIPIC_96",
            "PRE_PARENT_HIPIC_97","PRE_PARENT_HIPIC_98", "PRE_PARENT_HIPIC_99",
            "PRE_PARENT_HIPIC_101", "PRE_PARENT_HIPIC_102","PRE_PARENT_HIPIC_104",
            #Neuroticism
            "PRE_PARENT_HIPIC_113", "PRE_PARENT_HIPIC_115","PRE_PARENT_HIPIC_117",
            "PRE_PARENT_HIPIC_119", "PRE_PARENT_HIPIC_120",
             #Openness
            "PRE_PARENT_HIPIC_121", "PRE_PARENT_HIPIC_123","PRE_PARENT_HIPIC_139",
            "PRE_PARENT_HIPIC_143")

#Create dataframe with only variables that need to be recoded
attach(hipic_data)
table_r_vars<-subset(hipic_data,
                     select = var_names) 
detach(hipic_data)

#Recode values and assign new recoded variable name
for (i in 1:length(var_names)) {
  nam<-paste(var_names[i],"_r", sep="")
  assign(nam, recode_d(table_r_vars[i]))
}

# Combine Recoded variables into one dataframe
recoded_table<-data.frame(PRE_PARENT_HIPIC_1_r, PRE_PARENT_HIPIC_3_r,PRE_PARENT_HIPIC_4_r,
                     PRE_PARENT_HIPIC_11_r,PRE_PARENT_HIPIC_12_r, PRE_PARENT_HIPIC_29_r,
                     #Agreeableness
                     PRE_PARENT_HIPIC_34_r, PRE_PARENT_HIPIC_39_r,PRE_PARENT_HIPIC_45_r,
                     PRE_PARENT_HIPIC_53_r,PRE_PARENT_HIPIC_55_r, PRE_PARENT_HIPIC_56_r,
                     PRE_PARENT_HIPIC_57_r, PRE_PARENT_HIPIC_63_r,
                     #Conscientiousness
                     PRE_PARENT_HIPIC_80_r, PRE_PARENT_HIPIC_81_r,PRE_PARENT_HIPIC_83_r,
                     PRE_PARENT_HIPIC_84_r,PRE_PARENT_HIPIC_85_r, PRE_PARENT_HIPIC_91_r,
                     PRE_PARENT_HIPIC_94_r, PRE_PARENT_HIPIC_95_r,PRE_PARENT_HIPIC_96_r,
                     PRE_PARENT_HIPIC_97_r,PRE_PARENT_HIPIC_98_r, PRE_PARENT_HIPIC_99_r,
                     PRE_PARENT_HIPIC_101_r, PRE_PARENT_HIPIC_102_r,PRE_PARENT_HIPIC_104_r,
                     #Neuroticism
                     PRE_PARENT_HIPIC_113_r, PRE_PARENT_HIPIC_115_r, PRE_PARENT_HIPIC_117_r,
                     PRE_PARENT_HIPIC_119_r, PRE_PARENT_HIPIC_120_r,
                     #Openness
                     PRE_PARENT_HIPIC_121_r, PRE_PARENT_HIPIC_123_r,PRE_PARENT_HIPIC_139_r,
                     PRE_PARENT_HIPIC_143_r)

#Update Column Variable Names
colnames(recoded_table)<-c("PRE_PARENT_HIPIC_1_r", "PRE_PARENT_HIPIC_3_r","PRE_PARENT_HIPIC_4_r",
                          "PRE_PARENT_HIPIC_11_r","PRE_PARENT_HIPIC_12_r", "PRE_PARENT_HIPIC_29_r",
                          #Agreeableness
                          "PRE_PARENT_HIPIC_34_r", "PRE_PARENT_HIPIC_39_r","PRE_PARENT_HIPIC_45_r",
                          "PRE_PARENT_HIPIC_53_r","PRE_PARENT_HIPIC_55_r", "PRE_PARENT_HIPIC_56_r",
                          "PRE_PARENT_HIPIC_57_r", "PRE_PARENT_HIPIC_63_r",
                          #Conscientiousness
                          "PRE_PARENT_HIPIC_80_r", "PRE_PARENT_HIPIC_81_r","PRE_PARENT_HIPIC_83_r",
                          "PRE_PARENT_HIPIC_84_r","PRE_PARENT_HIPIC_85_r", "PRE_PARENT_HIPIC_91_r",
                          "PRE_PARENT_HIPIC_94_r", "PRE_PARENT_HIPIC_95_r","PRE_PARENT_HIPIC_96_r",
                          "PRE_PARENT_HIPIC_97_r","PRE_PARENT_HIPIC_98_r", "PRE_PARENT_HIPIC_99_r",
                          "PRE_PARENT_HIPIC_101_r", "PRE_PARENT_HIPIC_102_r","PRE_PARENT_HIPIC_104_r",
                          #Neuroticism
                          "PRE_PARENT_HIPIC_113_r", "PRE_PARENT_HIPIC_115_r","PRE_PARENT_HIPIC_117_r",
                          "PRE_PARENT_HIPIC_119_r", "PRE_PARENT_HIPIC_120_r",
                          #Openness
                          "PRE_PARENT_HIPIC_121_r", "PRE_PARENT_HIPIC_123_r","PRE_PARENT_HIPIC_139_r",
                          "PRE_PARENT_HIPIC_143_r")


#Append Recoded Variables into the hipic_data
hipic_datav2<- data.frame(hipic_data, recoded_table)

hipic_datav2<-subset(hipic_datav2,
               select = -c(PRE_PARENT_HIPIC_1,
                 PRE_PARENT_HIPIC_3,PRE_PARENT_HIPIC_4,
                 PRE_PARENT_HIPIC_11,PRE_PARENT_HIPIC_12, PRE_PARENT_HIPIC_29,
                 #Agreeableness
                 PRE_PARENT_HIPIC_34, PRE_PARENT_HIPIC_39,PRE_PARENT_HIPIC_45,
                 PRE_PARENT_HIPIC_53,PRE_PARENT_HIPIC_55, PRE_PARENT_HIPIC_56,
                 PRE_PARENT_HIPIC_57, PRE_PARENT_HIPIC_63,
                 #Conscientiousness
                 PRE_PARENT_HIPIC_80, PRE_PARENT_HIPIC_81,PRE_PARENT_HIPIC_83,
                 PRE_PARENT_HIPIC_84,PRE_PARENT_HIPIC_85, PRE_PARENT_HIPIC_91,
                 PRE_PARENT_HIPIC_94, PRE_PARENT_HIPIC_95,PRE_PARENT_HIPIC_96,
                 PRE_PARENT_HIPIC_97,PRE_PARENT_HIPIC_98, PRE_PARENT_HIPIC_99,
                 PRE_PARENT_HIPIC_101,PRE_PARENT_HIPIC_102,PRE_PARENT_HIPIC_104,
                 #Neuroticism
                 PRE_PARENT_HIPIC_113, PRE_PARENT_HIPIC_115,PRE_PARENT_HIPIC_117,
                 PRE_PARENT_HIPIC_119, PRE_PARENT_HIPIC_120,
                 #Openness
                 PRE_PARENT_HIPIC_121, PRE_PARENT_HIPIC_123,PRE_PARENT_HIPIC_139,
                 PRE_PARENT_HIPIC_143))
 
  

#check<-data.frame(hipic_datav2$PRE_PARENT_HIPIC_1, hipic_datav2$PRE_PARENT_HIPIC_1_r)

rm(list=setdiff(ls(), c("hipic_data", "hipic_datav2", "recode_d"))) 

#########
# Check alpha values
#########

#Original data
cronbach.alpha(hipic_data, na.rm = TRUE)

#Extraversion
cronbach.alpha(hipic_data[1:32], na.rm = TRUE)

#Agreeableness
cronbach.alpha(hipic_data[33:72], na.rm = TRUE)

#Conscientiousness
cronbach.alpha(hipic_data[73:104], na.rm = TRUE)

#Neuroticism
cronbach.alpha(hipic_data[105:120], na.rm = TRUE)

#Openess
cronbach.alpha(hipic_data[121:144], na.rm = TRUE)


#Recoded data - overall
cronbach.alpha(hipic_datav2, na.rm = TRUE)
#alpha value is 0.862

#Recoded data - Extaversion

grep("PRE_PARENT_HIPIC_2", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_32", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_1_r", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_29_r", names(hipic_datav2))

#Create Extraversion Data subset

ext_recode<-data.frame(hipic_datav2[1:26], hipic_datav2[107:112])

#Run cronbach alpha
cronbach.alpha(ext_recode,na.rm=TRUE)
  
#Recoded data - Agreeableness
grep("PRE_PARENT_HIPIC_33", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_72", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_34_r", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_63_r", names(hipic_datav2))

#Create agreeabless Data subset
agr_recode<-data.frame(hipic_datav2[27:58], hipic_datav2[113:120])

#Run cronbach alpha
cronbach.alpha(agr_recode,na.rm=TRUE)

#Recoded data - conscientiousness
grep("PRE_PARENT_HIPIC_73", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_103", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_80_r", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_104_r", names(hipic_datav2))

#Create conscientiousness Data subset
con_recode<-data.frame(hipic_datav2[59:75], hipic_datav2[121:135])

#Run cronbach alpha
cronbach.alpha(con_recode,na.rm=TRUE)

#Recoded data - Neuroticism
grep("PRE_PARENT_HIPIC_105", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_118", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_113_r", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_120_r", names(hipic_datav2))

#Create Neuroticism Data subset
neu_recode<-data.frame(hipic_datav2[76:86], hipic_datav2[136:140])

#Run cronbach alpha
cronbach.alpha(neu_recode,na.rm=TRUE)

#Recoded data - Openness
grep("PRE_PARENT_HIPIC_122", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_144", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_121_r", names(hipic_datav2))
grep("PRE_PARENT_HIPIC_143_r", names(hipic_datav2))

#Create Neuroticism Data subset
open_recode<-data.frame(hipic_datav2[87:106], hipic_datav2[141:144])

#Run cronbach alpha
cronbach.alpha(open_recode,na.rm=TRUE)


################################
#Correlation Matrices
################################
#Con
con_cor<- polychoric(con_recode)

rho.con <- con_cor$rho

```

## Background

- Autism (ASD) a neurodevelopmental disorder that affects ~ 1% of the population (APA, 2013)
- Children with autism may face challenges in responding effectively to
emotional situations and everyday encounters

- Studies found differences in personality traits between typically developing children and children with autism;Researchers found that typically developing children experienced higher levels of conscientiousness as compared to individuals with autism (Fortenberry et al., 2011).
  - Individuals with ASD are thought to show numerous unique personality
tendencies but that “surprisingly little research has examined these issues” (Scribner et al., 2014, p.1)

## Study Purpose

Goal of our exploratory study was to see whether the Hierarchical Personality Inventory for Children (HiPIC) questionnaire, which is normed on a general population, can be applied to children with Autism Spectrum Disorder (ASD).

## Hypothesis & Research Question

- The Big Five domains of personality exist within the autistic population but that the subdomains may differ since personality differences can affect individuals with ASD.
- Which personality traits emerge within the Conscientiousness domain (i.e., on the Big Five domains of personality) from the HiPIC questionnaire for this sample of autistic children?

## Data - Wood et al. (2019) Study 

- 195 participants, 7-13 years of age (*M* = 9.98, *SD* = 1.81)
- Recruited participants from 3 major US Universities through flyers, clinician referrals, and letters.
- Eligibility criteria:
  - IQ > 70
  - An ASD clinical diagnosis by 2 clinicians
  - Maladaptive and interfering anxiety
- 20% Female; 22% Latinx, 6% African American, 8% Asian/Pacific Islander, 76% White, 1% Native American or Alaskan, and 3% multiracial.

## Hierarchical Personality Inventory for Children (HiPIC)

- 144 item questionnaire completed by child's parent/caregiver 
- Measure the Big Five personality domains 
  - Extraversion, Agreeableness, Conscientiousness, Neuroticism, Openness
  - 18 subdomains split within the Big Five domains
- Instructed parents/caregivers to describe their child by their most frequent behaviors that occurred within the last year
- 5-point Likert scale (*barely characteristic* to *highly characteristic*)
- Focused on Conscientiousness
  - 32 items, 4 subdomains (Achievement, Orderliness, Concentration, and Persistence)

## Procedure

- Observed and reversed HiPIC Conscientiousness items to ensure consistent direction
- Confirmed with cronbach's $\alpha$ among cleaned conscientiousness items ($\alpha$ = 0.885)

## Procedure - Parallel Analysis

```{r echo=FALSE, results='hide'}
fa.parallel(rho.con, n.obs=195, fm="ml", fa="fa")
#Literature: 4 factors
#Parallel Analysis: 7 factors
#Eigenvalues above One: 4 factors
```

- Reveal 7 Factors, 4 factors with $\lambda>1$.


## Procedure - EFA & CFA Analysis

- Parallel analysis showed more factors than HiPIC was intended to measure
- Proceed with EFA:
  1. Conduct EFA and observe model fit indices
  2. Remove items that did not fit and/or had loadings $<0.45$ in all factors
  3. Repeat steps 1 & 2 until satisfactory model fit
- Proceed with CFA
  
## EFA Results

Table 1: EFA Model Results and Actions Taken

|EFA Analysis| TLI | RMSEA | RMSR  |Action  Taken              |
|------------|-----|-------|-------|---------------------------|
|7-Factor v1 |.731 |.099   | .040  | Removed 9 items           |      
|7-Factor v2 |.829 |.094   | .094  | Decide to run 6-Factor EFA|
|6-Factor v1 |.804 |.101   | .030  | Removed 1 item            |
|6-Factor v2 |.840 |.094   | .030  | Removed 1 item            |
|6-Factor v3 |.830 |.094   | .030  | Decide to run 5-Factor EFA|
|5-Factor    |.784 |.110   | .040  | Decide to run 4-Factor EFA|
|4-Factor    |.716 |.121   | .050  | Decide to run 6-Factor CFA|

## CFA Results

- 5-factor and 4-factor EFAs indicated less promising fit compared to the 6-factor model
- Ran a CFA for the 6-factor to determine fit of the conscientiousness items
  - TLI = 0.827
  - RMSEA = 0.098
  - SRMR = 0.073 
  - CFI = 0.856
  
## Discussion 

- "If you’ve met one person with autism, then you’ve met one person with autism." - *Professor Stephen Shore* 
- Great variation exists among individuals with ASD
- Relates with our study of conscientiousness
  - 4 factors in general population vs. 6 factors in our sample of children with ASD
  - achievement, orderliness, concentration, persistence, *organization*, *withdrawing*

## Limitations & Future Directions
- None of our fit indices fell into the acceptable range
  - Many were close; may have better fit with a larger sample
- Redesign a personality measure specifically for those with autism
- Include more items related to additional factors than suggested in prior
research

## References
 
- American Psychological Association. (2013). *Diagnostic and Statistical Manual of Mental Disorders* (DSM-5). American Psychiatric Association: DC.
- Fortenberry, C. L., Grist, C. L., & McCord, D. M. (2011). Personality trait differences between typically developing children and those diagnosed with autism spectrum disorder. *Individual Differences Research, 9,* 73-83.
- Jeste, S. S., & Geschwind, D. H. (2014). Disentangling the heterogeneity of autism spectrum disorder through genetic findings. *Nature reviews. Neurology, 10*(2), 74–81. https://doi.org/10.1038/nrneurol.2013.278.
- Mervielde, I., & De Fruyt, F. (1999). The construction of the hierarchical personality inventory for children (HiPIC). In I. Mervielde, I. Deary, F. De Fruyt, & F. Ostendorf (Eds.), *Personality Psychology* in Europe. Proceedings of the Eight European Conference on Personality Psychology (pp. 107-127). Tilburg: Tilburg University Press.

## References
- Schriber, R. A., Robins, R. W., & Solomon, M. (2014). Personality and self-insight in individuals with autism spectrum disorder. *Journal of personality and social psychology, 106*(1), 112.
- Schwartzman, B. C., Wood, J. J., & Kapp, S. K. (2016). Can the Five Factor Model of Personality Account for the Variability of Autism Symptom Expression? Multivariate Approaches to Behavioral Phenotyping in Adult Autism Spectrum Disorder. *Journal of autism and developmental disorders, 46*(1), 253–272. https://doi.org/10.1007/s10803-015-2571-x.
- Tackett, J. L., Kushner, S. C., De Fruyt, F., & Mervielde, I. (2013). Delineating personality traits in childhood and adolescence: Associations across measures, temperament, and behavioral problems. *Assessment*, 20(6), 738-751.

## References
- Wood, J.J., Kendall, P.C., Wood, K.S., Kerns, C.M., Seltzer, M., Small, B.J., Lewin, A.B., & Storch, E.A. (2019). Cognitive behavioral treatments for anxiety in children with autism spectrum disorder: A randomized clinical trial. *JAMA Psychiatry*. doi:10.1001/jamapsychiatry.2019.4160.  
